---
title: "shiny-example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{shiny-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```


## Using readMVT in Leaflet under Shiny

### Introduction

Although Leaflet for R can display very large image or raster data seamlessly with little effort it is far more limited in displaying vector data. The built-in methods require that your datasets be small enough to fit in the client's browser memory, and datasets of more than a very few thousand points are prohibitively slow to load.

One nice approach to displaying vector data is via [Mapbox Vector Tiles](https://docs.mapbox.com/data/tilesets/guides/vector-tiles-introduction/) (MVT) served by a [GeoServer](https://geoserver.org/). Tiling is done transparently on the GeoServer, so data prep doesn't require a lot of effort. The readMVT package facilitates fetching and displaying MVTs in Leaflet. It is intended to be run under Shiny, but can be used in simpler situations too.

This example displays vector stream data and road-stream crossings for Massachusetts, USA. It doesn't display the vector data until the user has zoomed to a specified level, for two reasons. First, there's no value in seeing points representing some 30,000 culverts and bridges when viewing the entire state. And second, loading large tiles at lower zoom levels would be prohibitively slow. When the user reaches the trigger zoom level, streams and crossings are displayed. They are continually loaded as the user zooms and pans above the trigger zoom. When the user zooms below the trigger level, the vector data are dropped from the display.

To speed things up considerably, data are cached at two levels. First, we use the package *memoise* to cache tiles as they are fetched from the GeoServer. If there are multiple client browsers under the same R session on a Shiny server, this cache is shared among users, so the same tile is never fetched from the GeoServer twice by the same R session. Second, Leaflet internally caches data drawn by the add\* functions (here, [leaflet::addPolylines()] and [leaflet::addCircleMarkers()]). We track (by user, not session) which tiles have been drawn so they don't need to be redrawn for a user.

If you want to try running this example, we've provided it in one big chunk at the bottom, all ready to be copied and pasted into a script.

### Setup

We start with some setup. First of all, we load the required packages.

```{r libraries}
library(readMVT)
library(shiny)
library(leaflet)
```

Now we set our home centerpoint and zoom level. The chosen values show all of Massachusetts.

```{r home}
home <- c(-71.6995, 42.1349)  # center of Massachusetts
zoom <- 8                     # starting zoom level (shows all of Massachusetts)
```

The next three variables have to do with zoom levels. MVTs are simplified at lower zoom levels. As we've chosen not to display vector data at lower zooms, we pick a fixed zoom to simplify caching and reduce fetching redundant data. At zoom level 14, data are not simplified, so that seems like a good choice.

```{r datazoom}
data.zoom <- 14   # all MVT tiles are read at this zoom level to simplify caching
```

Our trigger level is the zoom level to start showing vector data.

```{r trigger}
trigger <- 14     # show vector data when zoomed in this far or more
```

This variable tells Leaflet the zoom levels at which to display the vector data.

```{r zoomlevels}         
zoom.levels = 14:22    # show vector data at these zoom levels
```

Finally, we can use functions from readMVT. [read.XML()] fetches the GeoServer's "capabilities" file, an XML with information on all data on the GeoServer. Then, for each feature we will fetch, [layer.info()] extracts an MVT object from the XML. Here, we get layer info for stream linework and crossing (culverts and bridges) points.

```{r xml}         
xml <- read.XML('https://umassdsl.webgis1.com/geoserver')   # get capabilities of our GeoServer
streamlines <- layer.info(xml, 'testbed:streamlines')       # get info for stream linework
culverts <- layer.info(xml, 'testbed:CL_crossings7')        # get info for crossing points
```

Here's the MVT object for streamlines. It includes the layer name (*layer*), the bounding box (*box*), the tile boundaries at each zoom level (*tiles*), and a template URL that specifies the tile to fetch, which unresolved information (zoom, row, and column) enclosed in braces. In normal use of readMVT, you don't need to worry about these details--you just have to pass the result of [layer.info()] to [read.tile()] or [read.viewport.tiles()].

```{r streamlines}
> streamlines
$layer
[1] "testbed_streamlines"

$box
[1] -73.58043  41.22166 -69.89367  42.95567

$tiles
   zoom    rowmin    rowmax    colmin    colmax
1     0         0         0         0         0
2     1         0         0         0         0
3     2         1         1         1         1
4     3         2         2         2         2
5     4         5         5         4         4
6     5        11        11         9         9
7     6        23        23        18        19
8     7        47        47        37        39
9     8        94        95        75        78
10    9       188       191       151       156
11   10       376       383       302       313
12   11       752       766       605       626
  ...
31   30 394726401 401696265 317408725 328404932

$url
[1] https://umassdsl.webgis1.com/geoserver/gwc/service/wmts/rest/testbed:streamlines/simple_streams/EPSG:900913/EPSG:900913:{zoom}/{TileRow}/{TileCol}?format=application/vnd.mapbox-vector-tile
```

### User interface

Here's a simple Shiny user interface. We make a sidebar on the left to display the zoom level.

```{r ui}         
ui <- fluidPage(
   fluidRow(
      column(2,
             wellPanel(
                uiOutput('dashboard')
             )
      ),
      column(10,
             leafletOutput('map', height = '80vh')
      ))
)
```

### Server

Now we're ready for the guts of the example.

```{r server}     
server <- function(input, output, session) {
```

We display the Leaflet basemap and set the initial viewport. The ESRI World Street Map makes a good basemap for this example, as the streams match our vector streams.

```{r leaflet}         
output$map <- renderLeaflet({
   leaflet() |>
      addProviderTiles(providers$Esri.WorldStreetMap) |>
      setView(lng = home[1], lat = home[2], zoom = zoom)
})
```

Everything interesting happens inside an [leaflet::observe()]. First, we display the zoom level in the left sidebar, and "TRIGGERED!" if the zoom level is above the trigger. You should expect to see streams and crossings with TRIGGERED is shown.

```{r observe}    
observe({
   output$dashboard <- renderUI({
      output$dashboard <- renderUI({
         tagList(div('zoom = ', input$map_zoom,),
                 div(strong(if(isTRUE(input$map_zoom >= trigger)) ' TRIGGERED!')))
      })
   })
```

If we're zoomed out below the trigger zoom, we're done--the basemap will be displayed, but none of the vector data. If we're zoomed above (or equal to) the trigger, we'll read the vector tiles and render them. We use [leaflet::leafletProxy()] to modify the current display so there's no unnecessary (and ugly) redrawing.

```{r if}
if(isTRUE(input$map_zoom >= trigger)) {   # if above trigger zoom, draw features
   m <- leafletProxy('map', session)
```

We get the corner tiles of the current viewport at our fixed tile zoom level from Leaflet.

```{r viewport}         
nw <- get.tile(data.zoom, input$map_bounds$north, input$map_bounds$west)   # corners of viewport
se <- get.tile(data.zoom, input$map_bounds$south, input$map_bounds$east)
```

Now we read all vector tiles for streams in the viewport, omitting those we've already rendered. Rendering is captured by `session$userData[[layername]]`. Initially, it will be NULL, so `read.viewport.tiles` will initialize it. If you don't want to avoid redrawing, you can omit this argument and the following line that saves it. If there are any vector data in the viewport that haven't yet been rendered, `addPolylines` renders these streams.

```{r drawstreams}    
x <- read.viewport.tiles(streamlines, nw, se, data.zoom, session$userData[[streamlines$layer]])
session$userData[[streamlines$layer]] <- x$drawn
if(!is.null(x$tiles))
   m <- addPolylines(m, data = x$tiles, group = 'vector', opacity = 0.4, color = 'cornflowerblue', weight = 3)
```

We do the same thing for culverts. Note that we could move culverts into a separate if\-clause with a different trigger if we wanted to.

```{r drawculverts}
x <- read.viewport.tiles(culverts, nw, se, data.zoom, session$userData[[culverts$layer]])
session$userData[[culverts$layer]] <- x$drawn
if(!is.null(x$tiles))
   m <- addCircleMarkers(m, data = x$tiles, group = 'vector', opacity = 1, color = 'orange', radius = 4)
```

The call to [leaflet::groupOptions()]] displays all previously-rendered vector data at the selected zoom levels. Without this, zooming out would show streams and culverts for tiles we've visited, but not other visible tiles. That would be ugly. By hiding the vector data, they smoothly vanish as we zoom out. The Leaflet map is returned and rendered as the [shiny::observe()] finishes.

```{r finish}         
         return(groupOptions(m, 'vector', zoomLevels = zoom.levels))
      }
   })
}

shinyApp(ui, server)
```

## Example code, all in one place

If you're using RStudio, you should be able to install the named packages, copy this code, paste it into a script window, save it, and click "Run App."

```{r wholething} 
library(readMVT)
library(shiny)
library(leaflet)

home <- c(-71.6995, 42.1349)  # center of Massachusetts
zoom <- 8                     # starting zoom level (shows all of Massachusetts)

data.zoom <- 14               # all MVT tiles are read at this zoom level to simplify caching
trigger <- 14                 # show vector data when zoomed in this far or more
zoom.levels = 14:22           # show vector data at these zoom levels

xml <- read.XML('https://umassdsl.webgis1.com/geoserver')   # get capabilities of our GeoServer
streamlines <- layer.info(xml, 'testbed:streamlines')       # get info for stream linework
culverts <- layer.info(xml, 'testbed:CL_crossings7')        # get info for crossing points


# User interface ---------------------
ui <- fluidPage(
   fluidRow(
      column(2,
             wellPanel(
                uiOutput('dashboard')
             )
      ),
      column(10,
             leafletOutput('map', height = '80vh')
      ))
)


# Server -----------------------------
server <- function(input, output, session) {


   output$map <- renderLeaflet({
      leaflet() |>
         addProviderTiles(providers$Esri.WorldStreetMap) |>
         setView(lng = home[1], lat = home[2], zoom = zoom)
   })

   observe({
      output$dashboard <- renderUI({
         output$dashboard <- renderUI({
            tagList(div('zoom = ', input$map_zoom,),
                    div(strong(if(isTRUE(input$map_zoom >= trigger)) ' TRIGGERED!')))
         })
      })

      if(isTRUE(input$map_zoom >= trigger)) {   # if above trigger zoom, draw features
         m <- leafletProxy('map', session)
         
         nw <- get.tile(data.zoom, input$map_bounds$north, input$map_bounds$west)   # corners of viewport
         se <- get.tile(data.zoom, input$map_bounds$south, input$map_bounds$east)

         x <- read.viewport.tiles(streamlines, nw, se, data.zoom, session$userData[[streamlines$layer]])
         session$userData[[streamlines$layer]] <- x$drawn
         if(!is.null(x$tiles))
            m <- addPolylines(m, data = x$tiles, group = 'vector', opacity = 0.4, color = 'cornflowerblue', weight = 3)

         x <- read.viewport.tiles(culverts, nw, se, data.zoom, session$userData[[culverts$layer]])
         session$userData[[culverts$layer]] <- x$drawn
         if(!is.null(x$tiles))
            m <- addCircleMarkers(m, data = x$tiles, group = 'vector', opacity = 1, color = 'orange', radius = 4)

         return(groupOptions(m, 'vector', zoomLevels = zoom.levels))
      }
   })
}

shinyApp(ui, server)
```
